---
title: "Analysis of DEGs of those with ASD vs Controls"
author: "Matthew Dyer"
date: "2023-12-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Gather Data

The following chunk of code comes from the original author's work and is 
found at: https://www.ncbi.nlm.nih.gov/geo/geo2r/?acc=GSE28521
```{r}
#   Data plots for selected GEO samples
library(GEOquery)
library(limma)
library(umap)

# load series and platform data from GEO

gset <- getGEO("GSE28521", GSEMatrix =TRUE, getGPL=FALSE)
if (length(gset) > 1) idx <- grep("GPL6883", attr(gset, "names")) else idx <- 1
gset <- gset[[idx]]

ex <- exprs(gset)
# log2 transform
qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0), na.rm=T))
LogC <- (qx[5] > 100) ||
          (qx[6]-qx[1] > 50 && qx[2] > 0)
if (LogC) { ex[which(ex <= 0)] <- NaN
  ex <- log2(ex) }

# box-and-whisker plot
dev.new(width=3+ncol(gset)/6, height=5)
par(mar=c(7,4,2,1))
title <- paste ("GSE28521", "/", annotation(gset), sep ="")
boxplot(ex, boxwex=0.7, notch=T, main=title, outline=FALSE, las=2)
dev.off()

# expression value distribution plot
par(mar=c(4,4,2,1))
title <- paste ("GSE28521", "/", annotation(gset), " value distribution", sep ="")
plotDensities(ex, main=title, legend=F)

# mean-variance trend
ex <- na.omit(ex) # eliminate rows with NAs
plotSA(lmFit(ex), main="Mean variance trend, GSE28521")

# UMAP plot (multi-dimensional scaling)
ex <- ex[!duplicated(ex), ]  # remove duplicates
ump <- umap(t(ex), n_neighbors = 15, random_state = 123)
plot(ump$layout, main="UMAP plot, nbrs=15", xlab="", ylab="", pch=20, cex=1.5)
library("maptools")  # point labels without overlaps
pointLabel(ump$layout, labels = rownames(ump$layout), method="SANN", cex=0.6)
```

# Filter Data for only cortex tissue and establish factors for ASD and controls

```{r}
# Extract disease status information
disease_status <- gset$characteristics_ch1

# Create a logical vector (TRUE for autism, FALSE for control)
is_autism <- grepl("autism", disease_status, ignore.case = TRUE)

# Convert logical vector to numeric (1 for autism, 0 for control)
autism_numeric <- as.numeric(is_autism)

# Create a logical vector indicating whether each sample is from the cerebellum
is_cerebellum <- grepl("Cerebellum", gset$characteristics_ch1.1)

# Create a new vector excluding cerebellum samples
autism_numeric_filtered <- autism_numeric[!is_cerebellum]

# Subset the expression data to exclude cerebellum samples
data <- ex[, !is_cerebellum]
```

# Run SAMr

```{r, results='hide'}
library(samr)

#Split the data into autism and control groups
data.asd <- data[, autism_numeric_filtered == 1]
data.ctl <- data[, autism_numeric_filtered == 0]

# Combine the data
data.asd <- as.matrix(data.asd)
data.ctl <- as.matrix(data.ctl)
combined_data <- cbind(data.asd, data.ctl)

# Create a group vector with character values
group <- c(rep("ASD", ncol(data.asd)), rep("CTL", ncol(data.ctl)))

# Convert the group vector to a factor
group_factor <- as.factor(group)

# Convert group_factor to numeric values (2 for ASD and 1 for CTL)
y_numeric <- ifelse(group_factor == "CTL", 1, 2) # This makes ASD the point of comparison in regards to up and down regulation.

# Prepare the data list for SAM
sam.data <- list(
  x = combined_data,
  y = y_numeric,
  geneid = as.character(1:nrow(combined_data)),
  genenames = rownames(combined_data),
  logged2 = TRUE
)

# Run SAM
samr.obj <- samr(sam.data, resp.type="Two class unpaired", nperms=1000)

# Higher values are more stringent
# This delta value gives ~475 DEGs at a 0.0 "med false pos"   
delta = 1.444126120

delta.table <- samr.compute.delta.table(samr.obj) # Table used to find the optimal delta
siggenes.table <- samr.compute.siggenes.table(samr.obj, delta, sam.data, delta.table)
```

```{r}
up.genes <- siggenes.table$genes.up[,2] #121 genes
down.genes <- siggenes.table$genes.lo[,2] #354 genes
all.genes <- c(up.genes, down.genes) #475 genes total
up.genes.data <- combined_data[up.genes,]
down.genes.data <- combined_data[down.genes,]
f.data <- combined_data[all.genes,] #filtered data
```


# Hierarchical clustering
```{r}
library(dendextend)

# Calculate distance matrix
dist.mat.samples <- dist(t(f.data), method = 'euclidean')

# Perform hierarchical clustering
hclust.samples <- hclust(dist.mat.samples, method = 'ward.D2')

# Convert to dendrogram
dend <- as.dendrogram(hclust.samples)

# Define colors for each group
colors <- c(rep("blue", 29), rep("red", ncol(f.data) - 29))

# Apply the colors to the labels of the dendrogram
# Ensure the order of labels in dendrogram matches the column order in f.data
labels_colors <- setNames(colors, colnames(f.data))
dend <- color_labels(dend, col = labels_colors[labels(dend)])

# Plot the dendrogram
par(mar = c(5, 4, 2, 7))  # Adjust margins to fit legend
plot(dend, main = "Hierarchical Clustering of Samples")

# Add a legend
legend("topright", legend = c("ASD", "CTL"), fill = c("blue", "red"), cex = 0.8)
```

# Perform RF on the entire dataset and see what genes stand out. 

```{r}
library(randomForest)
combined.data <- t(combined_data)  # Transpose the data so that genes are columns. combined_data is the entire dataset w/9k genes
combined.data <- data.frame(group_factor, combined.data)

set.seed(123)  # for reproducibility
rf_model <- randomForest(group_factor ~ ., data = combined.data, ntree = 500, mtry = round(sqrt(ncol(combined.data)))) 
```

```{r}
rf_model #with this many genes (all 9163 genes), our Random Forest is only ~81.03% accurate
importance <- importance(rf_model)
top_genes <- head(importance[order(importance, decreasing = TRUE), ], 500)

rf.opt.data <- combined_data[names(top_genes), ] #subset with only the top 500 most important genes, reduce noise
combined.data.opt <- t(rf.opt.data)  # Transpose the data so that genes are columns
combined.data.opt <- data.frame(group_factor, combined.data.opt)

set.seed(123)  # for reproducibility
rf_model_opt <- randomForest(group_factor ~ ., data = combined.data.opt, ntree = 500, mtry = round(sqrt(ncol(combined_data)))) 
rf_model_opt #Increase to 86.21% accuracy with the top 500 genes that are relevant to the original RF. 
```

We can get even more accurate RF models, but I want to keep a larger gene pool for pathway analysis. 

```{r}
rf.names <- names(top_genes)
length(rf.names[!rf.names %in% down.genes]) #157/500 RF most important genes in the SAMr down regulated genes
length(rf.names[!rf.names %in% up.genes]) #56/500 RF most important genes in the SAMr up regulated genes

rf.sam.down <- rf.names[rf.names %in% down.genes]
rf.sam.up <- rf.names[rf.names %in% up.genes]
rf.sam.all <- c(rf.sam.down, rf.sam.up) #213 genes for pathway analysis
```

# Gene name conversions

```{r}
library(AnnotationDbi)
library(illuminaHumanv4.db)

gene.symbols.up <- select(illuminaHumanv4.db, 
                       keys = rf.sam.up, 
                       columns = c("SYMBOL"), 
                       keytype = "PROBEID")

gene.symbols.down <- select(illuminaHumanv4.db, 
                       keys = rf.sam.down, 
                       columns = c("SYMBOL"), 
                       keytype = "PROBEID")

score.up.genes <- data.frame(PROBEID = siggenes.table$genes.up[, "Gene ID"], 
                         Score_d = siggenes.table$genes.up[, "Score(d)"])

score.down.genes <- data.frame(PROBEID = siggenes.table$genes.lo[, "Gene ID"], 
                         Score_d = siggenes.table$genes.lo[, "Score(d)"])

gene.symbols.up <- merge(gene.symbols.up, score.up.genes, by = "PROBEID") 
gene.symbols.down <- merge(gene.symbols.down, score.down.genes, by = "PROBEID") 

combined.score.genes <- rbind(gene.symbols.up, gene.symbols.down)
```
# GSEA

```{r}
library(fgsea)
library(msigdbr)

# Function to perform fgsea for different MSigDB categories and subcategories
run_fgsea_for_categories <- function(gene_list) {
  # Define the categories and subcategories
  categories <- c("H", "C2", "C5")
  subcategories <- c("", "CP:KEGG", "CP:REACTOME", "GO:BP", "GO:CC", "GO:MF")

  # Initialize a list to store results
  fgsea_results <- list()

  # Iterate over categories and subcategories
  for (cat in categories) {
    for (subcat in subcategories) {
      # Construct the category string
      category_str <- ifelse(subcat == "", cat, paste(cat, subcat, sep = ":"))

      # Get MSigDB gene sets
      pathways <- tryCatch({
        msigdbr(species = "Homo sapiens", category = cat, subcategory = subcat)
      }, error = function(e) return(NULL))

      # Skip if the category/subcategory combination is not valid
      if (is.null(pathways)) next

      # Split gene sets
      gene_sets <- split(pathways$gene_symbol, pathways$gs_name)

      # Run fgsea
      fgsea_result <- fgsea(pathways = gene_sets, 
                            stats = gene_list, 
                            minSize = 5,   # Minimum size of gene set to consider
                            maxSize = 500)

      # Store results
      fgsea_results[[category_str]] <- fgsea_result
    }
  }

  return(fgsea_results)
}

# Create a named vector
gene.list <- setNames(as.numeric(combined.score.genes$Score_d), combined.score.genes$SYMBOL)

# Run the function
fgsea_all_results <- run_fgsea_for_categories(gene.list)

# Filter and save results
filtered_results <- lapply(fgsea_all_results, function(x) x[x$padj < 0.05, ])
```

# WGCNA

```{r}
library(WGCNA)

# Filter genes based on differential expression calculated from prior SAMr analysis and those identified by the optimized RF model. 
wgcna.ids <- combined.score.genes$PROBEID
wgcna.f.data <- f.data[rownames(f.data) %in% wgcna.ids, ]

dataExpr <- t(wgcna.f.data)

net = blockwiseModules(dataExpr, power = 7,
                        TOMType = "signed", minModuleSize = 10,
                        reassignThreshold =10, mergeCutHeight = 0.10,
                        numericLabels = TRUE, pamRespectsDendro = FALSE,
                        saveTOMs = TRUE,
                        saveTOMFileBase="TOM", verbose=3, ds=3)



```

# Look at modules

```{r}
mergedColors = labels2colors(net$colors)
plotDendroAndColors(net$dendrograms[[1]], mergedColors[net$blockGenes[[1]]],
"Module colors",
dendroLabels = FALSE, hang = 0.03,
addGuide = TRUE, guideHang = 0.05)
genes=colnames(dataExpr)
moduleColors=labels2colors(net$colors)
mymodules=cbind(genes,moduleColors)
```

```{r}
# Define the phenotype
# 1 for ASD individuals, 0 for controls, 
phenotype = c(rep(1, 29), rep(0, 29))

# Calculate Module Eigengenes
MEs0 = moduleEigengenes(dataExpr, moduleColors)$eigengenes
MEs = orderMEs(MEs0)

# Correlate Module Eigengenes with Phenotype
moduleTraitCor = cor(MEs, phenotype, use = "p")

# Degrees of freedom = number of samples - 2
moduleTraitPvalue = corPvalueStudent(moduleTraitCor, 22)
```

# Now let's visualize which modules are most significantly correlated with ASD

```{r}
# Convert p-values to -log10 scale for better visualization
negLogPvalues = -log10(moduleTraitPvalue)
moduleColorsVec = as.vector(rownames(moduleTraitCor))
moduleColorsVec = gsub("ME", "", moduleColorsVec)

# Plot the data
plot(moduleTraitCor, negLogPvalues, xlab = "Module-Trait Correlation", ylab = "-log10(p-value)", 
     main = "Module-Trait Correlation vs. P-value", pch = 19, col = moduleColorsVec)

# Add text labels for module names
text(moduleTraitCor, negLogPvalues, labels = names(moduleTraitCor), cex = 0.7, pos = 3)

# Add color labels next to each dot
offset = 0.1
for (i in 1:length(moduleTraitCor)) {
    text(moduleTraitCor[i], negLogPvalues[i] + offset, labels = moduleColorsVec[i], cex = 0.7, col = moduleColorsVec[i])
}

abline(h = -log10(0.05), col = "red", lty = 2)
```

# Export Network to Cytoscape

```{r}
load("C:/Users/Matt/Desktop/RBIF 114/Final/TOM-block.1.RData")

exportNetworkToCytoscape(
    adjMat = TOM,
    edgeFile = "C:/Users/Matt/Desktop/RBIF 114/Final/edgeFile.txt",
    nodeFile = "C:/Users/Matt/Desktop/RBIF 114/Final/nodeFile.txt",
    weighted = TRUE,
    threshold = 0.21, #have to use very low threshold to have enough genes for network analysis
    nodeNames = mymodules[,1],
    nodeAttr = mymodules[,2]
)
```

# Name conversions for later use

```{r}
finalnames <- rownames(f.data)

finalnames.final <- select(illuminaHumanv4.db, 
                       keys = finalnames, 
                       columns = c("SYMBOL"), 
                       keytype = "PROBEID")

# List of PROBEIDs for which you want to find Score(d)
probe_ids <- c("ILMN_1754727", "ILMN_1685496", "ILMN_1673704", "ILMN_1772627")

# Filter siggenes.table$genes.up to find matching PROBEIDs and extract Score(d)
score_d_values <- siggenes.table$genes.lo[siggenes.table$genes.lo[, "Gene ID"] %in% probe_ids, c("Gene ID", "Score(d)")]
colnames(score_d_values)[colnames(score_d_values) == "Gene ID"] <- "PROBEID"

final.results <- merge(score_d_values, finalnames.final, by = "PROBEID")
final.results
```

